---
title: "example1"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{example1}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(coranova)
library(ggplot2)
```

### Example 1: Comparing Three PGS in Two Populations

Say we are interested in comparing the performance of three polygenic scores in an African-ancestry and European-ancestry population, for example in the case of three polygenic scores generated with three different GWAS results: pgs1 from a multi-ancestry meta-analysis, pgs2 from a large European GWAS and pgs3 from a smaller-sized African GWAS. In this case, we are interested in answering a few questions:

#####  Q1. Is the PGS developed from the multi-ancestry meta-analysis better than the PGS developed with ancestry-specific GWASs? 
#####  Q2. Does the PGS developed from the multi-ancestry meta-analysis perform better in one population than the other?  

In this example we will demonstrate how to use our method to address these questions with the coranova command $\texttt{perform_coranova_parametric}$ . 

With coranova, we are comparing the performance of each PGS alone to predict the outcome using the correlation between the outcome and each PGS as our key metric ($cor(Y, PGS_i)$). 

### Data Preparation:
To use the coranova package, we need to have one dataframe per population sample where the rows are individuals in the population and the columns include the phenotype of interest, and the polygenic scores we wish to compare. 

After loading the package coranova, the two example data frames should be loaded, let's look at the first three rows using the `head` command.

We can see for each population data frame, we have a phenotype vector, and three polygenic scores, named `pgs1`, `pgs2`, and `pgs3`. 
```{r}
library(coranova)
head(afr, 3)
head(eur, 3)
```

Let's rename the polygenic score columns to match the GWAS with which they were built:
```{r}
colnames(afr) <- c("pheno", "META", "EUR", "AFR")
colnames(eur) <- c("pheno", "META", "EUR", "AFR")
```

### Visualizing Data

To further understand these data, we can also plot the correlations between the three polygenic scores and phenotype within the two populations:

```{r, fig.width = 6, fig.height = 4}
library(ggplot2)

cor_dat <- data.frame(pop = c(rep("AFR", 3), rep("EUR", 3)), 
                      cor = c(cor(afr)["pheno", c("META", "EUR", "AFR")],
                              cor(eur)["pheno", c("META", "EUR", "AFR")]), 
                      score = rep(c("META", "EUR", "AFR"), 2))

ggplot(cor_dat) + 
  geom_bar(aes(x = score, y = cor), stat = "identity") + 
  facet_wrap(~pop) + 
  labs(x = "Polygenic Score", y = "Correlation with Outcome")

```

Now, just looking at this figure we can see that the META polygenic score has a much higher correlation with the phenotype in both populations than the other two polygenic scores. However, we want to support this observation with statistical rigor, which we can do using the coranova package! We can also see that the correlation of the META polygenic score with the phenotype is higher in the European population sample than the African population sample, however we don't know if this difference is statistically significant. We can use this package to determine that!

### Using Coranova package
To use the coranova package,  we need to wrap the two population sample dataframes in a list with `list(afr, eur)` and provide the column names of the outcome (here `pheno`) and the names of the polygenic scores we want to compare (here: `META`, `EUR`, and `AFR`).

We will run the basic coranova command like this:
```{r, eval = F}
perform_coranova_parametric(list(afr, eur), "pheno", c("META", "AFR", "EUR"))
```


#### Comparing Scores across Populations

We first may want to establish whether a difference across the three scores within the populations exists, and whether we see a significant difference in score performance across the populations. We can use the basic coranova function \textt{perform_coranova_parametric} to do that.

When we run this:
```{r}
perform_coranova_parametric(list(afr, eur), "pheno", c("META", "AFR", "EUR"))
```

The first value `pB` is the p-value from the between test which evaluates whether the mean correlation of the three polygenic scores with the phenotype differs between the population groups `afr` and `eur`. Here the p-value is 0.0005, so we can conclude that the correlation of the scores do vary by population.

The second value `pW` is the p-value from the within test which evaluates whether the polygenic scores have the same correlation with the phenotype or not within the groups. Here the p-value is 9e-60,  so we can conclude that the scores do not perform the same in the two population samples.

The third value `pI` is the p-value from the interaction test which evaluates whether there is a difference in the pattern of score performance across the two groups. Here the p-value is 0.86 so we fail to reject the null hypothesis, we can also confirm this visually by noting that in both the `afr` and `eur` population samples, the best performing score is the `META` pgs, followed by the `EUR` pgs, followed by the `AFR` pgs.

#### Comparing Scores Within Each Population
To compare the scores further within each population, we can use $\texttt{perform_coranova_parametric}$ with a single population, and the scores we want to compare.

##### If we want to compare the performance of the three polygenic scores within each population,  we can use the following commands:
```{r}
perform_coranova_parametric(list(afr), "pheno", c("META", "AFR", "EUR"))
perform_coranova_parametric(list(eur), "pheno", c("META", "AFR", "EUR"))
```

However, since there are three polygenic scores, the test will just provide the within test p-value ($p_{within}$). If we want pairwise comparisons,  we can run the $\texttt{perform_coranova_parametric}$  with a single population and two polygenic scores. However, we recommend using adjusted p-value thresholds, due to the potentially high number of hypothesis tests when comparing pairwise. In this example, we have 2 population samples and three scores, so we will run 2*3 = 6 pairwise tests, thus we will use a p-value threshold of 0.05/6 = 0.0083. 

##### Comparing the pgs1 to the pgs2 to in each population sample:
```{r}
perform_coranova_parametric(list(afr), "pheno", c("META", "EUR"))
perform_coranova_parametric(list(eur), "pheno", c("META", "EUR"))
```
When we only include two polygenic scores, the function returns the difference between the two scores, the standard error of the difference, and its 95% confidence interval.

We conclude that in both population samples, the META has a higher correlation with the outcome than the EUR polygenic score. In the African population sample, the META PGS has 0.16 higher correlation with the outcome than the EUR PGS ($95\% \text{ CI:} (0.12, 0.19), p_{within} =  3.4e^{-20}$). In the European population sample, the META PGS has 0.15 higher correlation with the outcome than the EUR PGS ($95\% \text{ CI:} (0.1, 0.18), p_{within} =  3.0e^{-19}$). 

SIDENOTE: We can use the interaction test implemented in coranova to determine whether the difference in performance between the META PGS and EUR PGS differs across the two populations:
```{r}
perform_coranova_parametric(list(afr, eur), "pheno", c("META", "EUR"))
```
Here, $p_{interaction}$ is 0.59,and thus we conclude that the pattern of score performance does not differ across population samples.

##### We can also compare the META PGS to the AFR PGS to in each population sample:
```{r}
perform_coranova_parametric(list(afr), "pheno", c("META", "AFR"))
perform_coranova_parametric(list(eur), "pheno",  c("META", "AFR"))
```
We conclude that in both population samples, the META PGS has a higher correlation with the outcome than the AFRP PGS. 

##### Finally, we can also compare the EUR PGS to the AFR PGS to in each population sample:
```{r}
perform_coranova_parametric(list(afr), "pheno", c("EUR", "AFR"))
perform_coranova_parametric(list(eur), "pheno", c("EUR", "AFR"))
```
Here we do not find a significant difference in performance between the EUR and AFR PGS in either population ($p_{within}$ = 0.3 in the African sample, $p_{within}$ = 0.1 in the European sample).

#### Question 4: Comparing META PGS Across Populations

Now that we have established that the META PGS performs best in each population,  we can compare its performance between populations using the "between" test. If given two populations and a single polygenic score, the function $\texttt{perform_coranova_parametric}$ will run this test automatically and provide the difference and 95% confidence interval.
```{r}
perform_coranova_parametric(list(eur, afr), "pheno", "META")
```
When we run this test, we find that the score has a 0.036 higher correlation with the outcome in the European population sample than the African population sample (95% CI: (0.002, 0.072)), and conclude that this difference is significant ($p_{between} = 0.04$). 

### Conclusions

In this example, we have utilized the coranova package and its command $\texttt{perform_coranova_parametric}$ to compare three polygenic scores in two population samples. 

We established that the scores' correlation with the outcome did substantially vary within and across populations. We then considered the pairwise differences between the scores to determine than in both populations the `META` PGS has a statistically significantly higher correlation with the outcome than the other two ancestry-specific scores. Finally, we found that the `META` PGS has a slightly higher  correlation with the outcome in the European population sample compared to the African population sample, and this difference is statistically significant.

In the second example, we will see how we can use another command in the coranova package $\texttt{perform_alt_test}$ to devise more complicated hypothesis tests! 


