---
title: "example2"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{example2}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.height = 4, fig.width = 6
  
)
```

```{r setup}
library(coranova)
library(dplyr)
library(ggplot2)
library(Matrix)
```

In this example, we will demonstrate how to use the function $\texttt{perform_alt_test}$ to define and test additional hypotheses!


### Some Theory to Set the Scene
First, let's go through a little bit of the math that makes this method possible.

The correlation between two random variables summarizes the strength of their linear relationship, a correlation close to 1 indicates that the two variables have a strong, positive correlation.

In the case of polygenic scores, we would like a PGS to have a high correlation with its associated outcome and we can use correlation to compare PGS.

To derive statistical tests of correlation measures, we can model correlation coefficients as normally distributed, as defined by Olkin & Finn (1990).

Let $u$ be a row vector of sample correlations of $p$ PGS with outcome $Y$ in population $j$ and let $\mu$ denote the vector of population values. \

$u = (r_{1, j}, r_{2,j}, ... r_{p,j})$\
$\mu = (\rho_{1, j}, \rho_{2,j}, ... \rho_{p,j})$\

We can model $\mu$ as normally distributed with covariance matrix $\hat{\Sigma}(\mu)$, as defined by Olkin & Finn (1990).

$$\mu \sim N(\mu_0 , \hat{\Sigma}(\mu)) $$

Then, we can test the null hypothesis,
$$H_0: \mu A = \mu_0$$
where $A$ is a $p$ x $M$ contrast matrix of rank $M$. Then, Then the test-statistic 
$$ S = (u A - \mu_0) (A\hat{\Sigma}_\infty(u) A')^{-1}(u A - \mu_0)'$$
will be $\chi^2$ distributed with degrees of freedom $M$ (the rank of contrast matrix $A$).

For example: 

Let $\mu_0 = 0 \text{ and } A^\intercal= \begin{bmatrix} 1 & -1 & 0 & ... & 0  \end{bmatrix}$

Then, 
$H_0: \rho_{1, j} = \rho_{2, j}$. 

and the null hypothesis is that the first polygenic score has the same correlation with the outcome as the second polygenic score in population j.

### Application to PGS

Now let's apply this thinking to an example.

In this example, we will consider 12 PGS prepared using four GWAS bases and three methods (so 4 x 3 total PGS) across two populations. 

Let's take a peek at the data:
```{r}
head(pop1, 3)
head(pop2, 3)
```

and look at correlation with the simulated outcome among the scores:
```{r, fig.height=4, fig.width=6}
cor_dat <- data.frame(pop = c(rep("pop1", 12), rep("pop2",12)),
                      method = rep(c(rep("A", 4), rep("B", 4), rep("C", 4)), 2),
                      cor = c(cor(pop1)["pheno", colnames(pop1)[-1]],
                              cor(pop2)["pheno", colnames(pop2)[-1]]) ,
                      gwas = rep(c("gwas1", "gwas2", "gwas3","gwas4"), 6))
```

In this example, we might be more concerned with comparing methods:

```{r}
cor_dat %>% mutate(score = paste(method, gwas, sep = "_")) %>%
  ggplot() + geom_bar(aes(x = method, y = cor), stat = "identity") + 
  facet_grid(gwas~pop) + 
  labs(x = "Polygenic Score Method", y = "Correlation with Outcome")

```

or we might be more concerned with comparing GWAS bases:
```{r}
cor_dat %>% mutate(score = paste(method, gwas, sep = "_")) %>%
  ggplot() + geom_bar(aes(x = gwas, y = cor), stat = "identity") + 
  facet_grid(method~pop) + 
  labs(x = "Polygenic Score Base GWAS", y = "Correlation with Outcome")
```

There are many questions we may want to investigate with these data!

#### Let's start with the small case of comparing the performance of the three methods with gwas1 as the base. 
```{r}
cor_dat %>% mutate(score = paste(method, gwas, sep = "_")) %>%
  filter(gwas == "gwas1") %>% 
  ggplot() + geom_bar(aes(x = method, y = cor), stat = "identity") + 
  facet_grid(~pop) + 
  labs(x = "Polygenic Score Base GWAS", y = "Correlation with Outcome")
```


We can use the function $\texttt{perform_alt_test}$, to implement this type of hypothesis testing, we just need to specify the:\

`dat_list`: list of data frames, where each data frame refers to a separate population sample

`outcome`: name of outcome variable (must be common across dataframes in dat_list)

`measures`: vector of the names of the columns of the measures to be compared

`contrast`: contrast matrix to generate hypothesis test

`method`: to request parametric ("parametric") or bootstrap implementation ("boot") of covariance matrix V

`B`: number of bootstraps if method == "boot" is chosen

In this example, we'll demonstrate how we can define a contrast matrix to determine whether there is a difference in score performance by methods. We will consider only pop1 at first, then show how we can extend this contrast matrix to two populations. 


First, let's define the key variables that the function needs:
```{r}
dat_list <- list(pop1)
outcome <- "pheno"
measures <- c("gwas1_methodA", "gwas1_methodB",  "gwas1_methodC")
method <- "parametric" 
```

Now, we need to define a contrast matrix to define our null hypothesis to be tested. 

Because we have only specified "gwas1_methodA", "gwas1_methodB", and "gwas1_methodC" as our measures, and are only considering pop1, our vector u is:

$$u = (r_{\text{gwas1_methodA, pop1}}, r_{\text{gwas1_methodB, pop1}}, r_{\text{gwas1_methodC, pop1}})$$

if we are interested in comparing the three methods built with the first gwas,
we can define a contrast matrix like so

 $$A = \begin{bmatrix} 1 & -1 & 0 \\
1 & 0 & -1  \end{bmatrix}$$
```{r}
contrast <- matrix(c(1, -1, 0,
                     1, 0, -1), nrow = 2, ncol = 3, byrow = T)
```


when we matrix multiply our contrast matrix by our vector of correlations ($u$),

$$u A = \begin{bmatrix} r_{\text{gwas1_methodA, pop1}} -  r_{\text{gwas1_methodB, pop1}}\\
r_{\text{gwas1_methodA, pop1}} -  r_{\text{gwas1_methodC, pop1}} \end{bmatrix} $$

we can see that our null hypothesis is:

$$u A = \begin{bmatrix} r_{\text{gwas1_methodA, pop1}} -  r_{\text{gwas1_methodB, pop1}}\\
r_{\text{gwas1_methodA, pop1}} -  r_{\text{gwas1_methodC, pop1}} \end{bmatrix}  = \begin{bmatrix} 0\\
0 \end{bmatrix} $$

and our alternative hypothesis is that at least one of the pairwise differences (either $r_{\text{gwas1_methodA, pop1}} -  r_{\text{gwas1_methodB, pop1}}$ or $r_{\text{gwas1_methodA, pop1}} -  r_{\text{gwas1_methodC, pop1}}$) is not equal to 0 . 

we can test this null  hypothesis and generate our test-statistic and p-value using the \texttt{perform_alt_test} function.
```{r}
perform_alt_test(list(pop1), outcome, measures, contrast, "parametric")
```

And finally, we conclude that at least one of the pairwise differences (either $r_{\text{gwas1_methodA, pop1}} -  r_{\text{gwas1_methodB, pop1}}$ or $r_{\text{gwas1_methodA, pop1}} -  r_{\text{gwas1_methodC, pop1}}$) is not equal to 0 ($p = 8e^{-11}$), indicating that there is a difference in the scores generated with the different methods built with gwas1. 

##### To understand how we got to this contrast matrix, let's explain what's going on under the hood of the function.

First, we can generate some of the objects used within the function:
```{r}
cormat_list <- lapply(dat_list, cor) #list of correlation matrices, one per 
                                     #population sample, here we have just 1 pop, 
                                     # so list of size 1
n_list <- lapply(dat_list, nrow) #list of sample sizes, again only 1 pop so, 
                                 # so list of size 1
```

To look at the correlation matrix for the polygenic scores we are considering in this example, we can subset like so:
```{r}
cormat_list[[1]][c(outcome, measures), c(outcome, measures)]
```

Now, let's populate the vector of correlations and their covariance matrix for our test statistic.
```{r}
u <- populate_u(cormat_list, outcome, measures) #u is our vector of the correlations between the polygenic scores (specified in the measures vector) and the outcome
u
V <- populate_sigma(cormat_list, outcome, measures, n_list) #V is the covariance matrix of R, we need this to derive our test statistic!
V
```


```{r}
#we can see that the first term is equal to the correlation of the pgs derived with method A and the outcome (R[1])
# minus the the correlation of the pgs derived with method A and the outcome (R[2])
#and the second term is equal to the correlation of the pgs derived with method A and the outcome (R[1])
# minus the the correlation of the pgs derived with method A and the outcome (R[3])
contrast %*% u
unname(u[1] - u[2])
unname(u[1] - u[3])
```

Within the function, the test statistic is derived as follows:
```{r}
S <- t(contrast %*% u) %*% solve(contrast %*% V %*% t(contrast)) %*% (contrast %*% u)


p <- pchisq(as.numeric(S), rankMatrix(contrast), lower.tail = F)
```

and this p-value is the probability that our test statistic is at least as extreme as our observed results given the null hypothesis.
```{r}
S
p
```

which is exactly the same as the results of \texttt{perform_alt_test}.
```{r}
perform_alt_test(list(pop1), outcome, measures, contrast, "parametric")
```

#### That was a lot! But now that we know what we're working with extending to two populations will be no problem!

To compare the methods across gwas 1 in both populations,

let $$u = (r_{\text{gwas1_methodA, pop1}}, r_{\text{gwas1_methodB, pop1}}, r_{\text{gwas1_methodC, pop1}}, \\r_{\text{gwas1_methodA, pop2}}, r_{\text{gwas1_methodB, pop2}}, r_{\text{gwas1_methodC, pop2}})$$

we can devise a contrast matrix like so:

 $$A = \begin{bmatrix} 1 & -1 & 0 &  1 & -1 & 0 \\
1 & 0 & -1 & 1 & 0 & -1  \end{bmatrix}$$

Then, our null hypothesis is:

$$u A = \begin{bmatrix} (r_{\text{gwas1_methodA, pop1}} + r_{\text{gwas1_methodA, pop2}}) -  (r_{\text{gwas1_methodB, pop1}} + r_{\text{gwas1_methodB, pop2}})\\
(r_{\text{gwas1_methodA, pop1}} + r_{\text{gwas1_methodA, pop2}}) -  (r_{\text{gwas1_methodC, pop1}} + r_{\text{gwas1_methodC, pop2}})\end{bmatrix}  = \begin{bmatrix} 0\\
0 \end{bmatrix} $$

Essentially here we are averaging across the populations!

Now that we know what we're doing, implementing this is a lot easier. All we need to do is provide the contrast matrix and plug it in to the $\texttt{perform_alt_test}$ function. Just make sure you have input the correct population/s, outcome and measures variables!
```{r}
contrast <- matrix(c(1, -1, 0, 1, -1, 0,
                     1, 0, -1, 1, 0, -1), nrow = 2, ncol = 6, byrow = T)

perform_alt_test(list(pop1, pop2), outcome, measures, contrast, "parametric")
```

#### Following the same logic, if we want to average across another gwas, all we need to do is change the "measures" option:

Here we compare across the scores built with gwas2:
```{r}
contrast <- matrix(c(1, -1, 0, 1, -1, 0,
                     1, 0, -1, 1, 0, -1), nrow = 2, ncol = 6, byrow = T)

measures <- c("gwas2_methodA", "gwas2_methodB",  "gwas2_methodC")
perform_alt_test(list(pop1, pop2), outcome, measures, contrast, "parametric")
```

We can also fully average across all the gwas options and the populations:
$$u A = \begin{bmatrix} \sum_{i = 1}^4 (r_{\text{gwas}_i\text{_methodA, pop1}} + r_{\text{gwas}_i\text{_methodA, pop2}}) -  (  r_{\text{gwas}_i\text{_methodB, pop1}} + r_{\text{gwas}_i\text{_methodB, pop2}})\\
\sum_{i = 1}^4 (r_{\text{gwas}_i\text{_methodA, pop1}} + r_{\text{gwas}_i\text{_methodA, pop2}}) -  (r_{\text{gwas}_i\text{_methodC, pop1}} + r_{\text{gwas}_i\text{_methodC, pop2}})\end{bmatrix}  = \begin{bmatrix} 0\\
0 \end{bmatrix} $$
```{r}
measures <- c("gwas1_methodA", "gwas1_methodB",  "gwas1_methodC",
              "gwas2_methodA", "gwas2_methodB",  "gwas2_methodC",
              "gwas3_methodA", "gwas3_methodB",  "gwas3_methodC",
              "gwas4_methodA", "gwas4_methodB",  "gwas4_methodC")

contrast <- matrix( c(rep(c(1, -1, 0, 1, -1, 0),4),
                     rep(c(1, 0, -1, 1, 0, -1), 4)), nrow = 2, ncol = 24, byrow = T)

perform_alt_test(list(pop1, pop2), outcome, measures, contrast, "parametric")
```

